 

### 2023-03-11 paper3.Rmd
Revisions throughout and final version for External Reader

paper3_abstract <- paste0("For searches related to depression, we observe a sustained decline in search density of ",
                          function_dynamic_text_results("main","depression", "slope" , TRUE, TRUE), " and an immediate downward level shift of ",
                          function_dynamic_text_results("main","depression", "intercept", TRUE),
                          " relative to comparable symptom searches. For anxiety, we estimated a similar slope change of",
                          function_dynamic_text_results("main","anxiety", "slope", TRUE), 
                          " and a level shift of ",
                          function_dynamic_text_results("main","anxiety", "intercept", TRUE),
 ". We did not observe significant variations across predominately Democrat or Republican voting states and counties, and our findings were robust to alternative specifications of bandwidth, fixed effects, and data cleaning decisions.")

@@

\emph{Daniel Arias, Jessica Cohen, Karestan Koenen, Margaret McConnell, Stéphane Verguet}

\vspace{24pt}

@@

{\bf Methods:} We utilized controlled interrupted time series approaches to evaluate whether the development of COVID-19 vaccines reduced Google searches related to symptoms of depression and anxiety in the U.S.. We examined how search trends changed following the announcement of vaccine safety and efficacy data for mental health symptoms compared with physical health symptoms, and estimated associations between vaccine announcement and changes in mental health symptom search density. We adjusted for state, month, and weekday effects, and applied numerous sensitivity analyses to assess the robustness of our findings. To explore heterogeneity by predominate political affiliation, we also tested whether our estimates varied among states and counties that predominantly voted for the Democratic Party candidate in the 2020 presidential election compared to the Republican candidate.

{\bf Findings and interpretation:} Under our main model specification, we observed reduced search density for depression and anxiety associated with the announcement of vaccine safety data. `r paper3_abstract` If symptom search density echoes population health, our results would indicate that announcements regarding the development of safe and effective COVID-19 vaccines led to a robust and sustained improvement in mental health. Our findings underscore the importance of timely and transparent communication and highlight the potential application of high frequency internet search data for population mental health surveillance.

@@

## Introduction
The COVID-19 pandemic has had a profound effect on the mental health and wellbeing of individuals worldwide. Studies from Australia,[@batterhamTrajectoriesDepressionAnxiety2021] China,[@ahmedEpidemicCOVID19China2020; @gaoMentalHealthProblems2020; @hwangLonelinessSocialIsolation; @leiComparisonPrevalenceAssociated2020; @liuPrevalencePredictorsPTSS2020; @qiuNationwideSurveyPsychological2020; @wangImmediatePsychologicalResponses2020; @wangStudyPublicPsychological2021; @wangPsychologicalDistressCoping2020; @zhangImpactCOVID19Pandemic2020] Denmark,[@sonderskovDepressiveStateDenmark] Europe,[@hajekPrevalenceDeterminantsProbable2022] Iran,[@moghanibashi-mansouriehAssessingAnxietyLevel2020] Ireland,[@hylandAnxietyDepressionRepublic2020]  Italy,[@mazzaNationwideSurveyPsychological2020; @mocciaAffectiveTemperamentAttachment2020] Nepal,[@samadarshiOnlineSurveyFactors2020] New Zealand,[@gasteigerDepressionAnxietyStress2021; @lawRiskFactorsChanges2022] Spain,[@gonzalez-sanguinoMentalHealthConsequences2020; @ozamiz-etxebarriaStressAnxietyDepression2020] Turkey[@ozdinLevelsPredictorsAnxiety2020],  the United States,[@jiaNationalStateTrends2021, p. 2020–2021; @kujawaExposureCOVID19Pandemic2020] and the United Kingdom[@dettmannInvestigatingPrevalenceAnxiety2022] have reported high levels of anxiety (6-51%), depression (15-48%), and post-traumatic stress disorder (7-54%) associated with the onset of the pandemic. In the United States, approximately four in ten adults reported experiencing symptoms of anxiety or depression in December 2020—a four-fold increase from January-June 2019.[@panchalImplicationsCOVID19Mental2021, p. 19] Globally, it has been estimated that the prevalence of depression and anxiety during the COVID-19 pandemic may be as high as 24% and 21%, respectively.[@banko-ferranImpactCOVID19Mental2020; @castaldelli-maiaInvestigatingEffectNational2021] This striking burden has been described as a mental health crisis[@melimopoulosHowCOVID19Pandemic2021] and as a pandemic in its own right.[@kwaiWhatPointYoung2021]

@@

While this literature suggests that receiving a COVID-19 vaccine can improve mental health, the potential mental health benefits of announcements regarding the development of vaccines have received comparatively less attention.[@cotfasLongestMonthAnalyzing2021; @kesselEffectCOVID19Pandemic2023] In the context of a deadly pandemic, the announcement of a safe and effective vaccine could be a powerful signal, providing a psychological boost that the worst might soon be mitigated. Studying how population mental health may have been impacted by the release of vaccine efficacy and safety trial data may provide insights into how emotional well-being changed throughout the pandemic in response to these signals, which, in turn, could inform public health communication and contribute to our understanding of how the public responds to information about vaccines—a necessary understanding for increasing vaccine uptake and trust in public health responses.

@@

Our study analyzes whether news announcements (in November 2020) regarding the imminent availability of safe and effective COVID-19 vaccines impacted population mental health—even before these vaccines were made available to the public. Using Google search density data, we applied interrupted time series approaches to investigate how these announcements might have affected search trends for mental health symptoms in the U.S. and five other countries where data were available: Australia, the United Kingdom, Ireland, New Zealand, Singapore. Owing to emerging evidence that political ideology may be a predictor of confidence in public health and vaccine hesitancy,[@debusPoliticalIdeologyVaccination2021; @baumgaertnerInfluencePoliticalIdeology2018; @stolerSociopoliticalPsychologicalCorrelates2022] we further explored whether changes in search trends for mental health symptoms varied across U.S. states and counties that predominantly voted Democrat vs. Republican in the 2020 presidential election. Our findings suggest declines in density for searches related to anxiety and depression in all of the countries examined; in the U.S., these declines were found to be consistent irrespective of political ideology.

@@

We take as our first credible signal of the imminent availability of safe and effective COVID-19 vaccines to be November 16, 2020. On November 9^th^, Pzifer announced that early vaccine trial data showed 90% efficacy[@thomasPfizerEarlyData2020] and on the 16^th^, Moderna announced preliminary results showing 95% efficacy.[@gradyEarlyDataShow2020] While U.S. Food and Drug Administration (FDA) Emergency Use Authorization (EUA) was not granted until December, the concurrent announcements of vaccine efficacy were widely reported and interpreted by many public health practitioners, policymakers, and the general public as a substantial signal of a new phase of the pandemic, with Moderna’s chief executive officer calling the news “a game changer” in the fight against COVID-19.

@@

We accessed aggregated national and subnational data from Google’s COVID-19 Search Trends Symptoms Dataset [@Wahltinez2020] (hereafter referred to as the symptoms dataset) between January 1, 2020 and April 1, 2021. The symptoms dataset provides time series data of search term density, standardized to pairs of health symptoms and regions (e.g., searches for depression in Australia or fever in Cook County, Illinois) and normalized within each pair based on the symptom’s relative popularity. The time series of search density for each symptom-region pair also contains artificial noise, as Google’s differential privacy technique to data aggregation and anonymization adds random noise to protect user privacy (the errors are symmetrically distributed).[@Wahltinez2020]

Search density data for over 400 symptoms were reported by Google across national and subnational regions of six predominately English-speaking countries: Australia, the United Kingdom, Ireland, New Zealand, Singapore, and the U.S. In the U.S., subnational regions include states and counties.  Because the aggregated search density data were constructed based on relative search population within each region, we could not compare search density for specific symptoms across different regions; however, comparing search density within regions (or a group of regions) was feasible both over time and across different symptoms.

@@

We utilized three levels of region: 1) national level for the six countries; 2) U.S. state level data across `r n_states` states and the District of Colombia; and 3) U.S. county level across `r n_counties` counties (though there are over 3,000 counties in the U.S., due to data privacy concerns, smaller county results were not included in the symptoms dataset). Following the convention of the symptoms dataset, we will refer to the geographic units (e.g., states, counties, and nations) within the data as "regions."

@@

We obtained county-level presidential election results for 2020 from the MIT Election Data and Science Lab,[@mitelectiondataandsciencelabCountyPresidentialElection2022] which we used to calculate aggregate county and state votes (including absentee, early, and election day ballots) for each candidate. The aggregate votes were then used to determine vote share and margins (e.g., the percentage points difference in vote share between the winning candidate and the runner up). Applying a similar approach as other research into COVID-19 and political party inclination,[@grossmannAllStatesClose2021] we used these margins to sort counties and states into quintiles, such that the 20% most Republican-voting localities appearing in the symptoms dataset were sorted into the first quintile and the 20% most Democratic-voting localities sorted into the fifth quintile.

@@

We used an interrupted time series (ITS) approach to model the Google search density for a given symptom $d$ using the following baseline equation: $$Y_{d,g, i} = \lambda_0 + \lambda_1 time_t + \alpha_d Post_{d,i} + \beta_d time_t*post_{d,i} + \theta_r + T^{'}_t\chi + \epsilon_{it} \ \ \ ,\ \ \ \ \ \ \ \ \ \ \ \ (3.1)$$ where $time_t$ is a continuous variable that captures the days before the vaccine announcement ($time_0$), $Post_i$ is an indicator variable equal to 1 in the period after the announcement (0 otherwise), $\theta_r$ indicates a series of fixed effects for each region $(r)$, $T^{'}_t\chi$ is a matrix of time fixed effects (i.e., month and day of week), and $\epsilon_{it}$ is an error term. $\alpha_d$ can be interpreted as a level change associated with the announcement date—that is, the immediate difference in search density for a disease $d$ following November 19, 2020; $\beta_d$—the slope change—can be interpreted as the sustained change in search density in the days following November 16th.

@@

```
We apply this model for all available `r diseases_n` diseases with county-level data. As search density may rapidly change in response to new announcements, we trimmed the bandwidth of dates to fit our model to 40 days before and after November 16, 2020.

To control for time-varying confounders which may affect symptom search trends, we applied a controlled interrupted time series (CITS) approach to contrast how searches for mental health symptoms varied above and beyond changes in search density for a comparable physical complaint (i.e., headaches).

After reviewing the list of symptoms for which search density data were available, we identified the following `r mh_diseases_n` mental health symptoms: `r paste0(gsub("asperger", "Asperger", gsub("_", " ", printList(toPrint = sort(list_mental_health_symptoms), finalSepWord = "and", midSep = ","))),".")`

For each of these symptoms, $d$, we ran the following model:

@@

Given the high number of observations in our sample, we evaluated significance against an alpha of 0.01, and we constructed 99% confidence intervals (CIs) using heteroskedasticity-robust standard errors.

@@

As with other quasi-experimental methods, it is recommended that the modeling specifications be evaluated using a sensitivity analysis. To do so, we defined a list of seven methodological choices concerning data processing and modeling assumptions and interacted these choices to generate all plausible combinations. These combinations were then conducted to allow for a multiverse analysis to identify how sensitive effect estimates and CIs were to different methodological choices. The criteria used to generate the plausible model combinations are outlined in Table 3.1.

@@

First, we considered the role of data cleaning in the outcome data, which involved outlier smoothing using Friedman’s nonparametric super smoother regression estimator.[@friedmanVariableSpanScatterplot1984] Second, we compared the level of region of aggregated data, using U.S. counties, U.S. states, and, finally, national trends from the six counties in the sample. Third, we evaluated sensitivity to our specification of the bandwidth cut-off around the announcement date, extending it from ±40 days to ±50 and ±60 days. Fourth, we toggled whether to account for day of week fixed effects; fifth, we toggled month fixed effects; and sixth, we toggled regional fixed effects (i.e., fixed effects for states in the U.S. state level data, fixed effects for counties in the U.S. county level data, etc.).

@@

The first subgroup analysis applied the CITS approach against all `r mh_diseases_n` mental health related symptoms against search data from each of the six countries in the sample, using our nationally aggregated data. This analysis aimed to investigate the extent to which effects observed in U.S. states and counties were also seen in other countries.

The second subgroup analysis applied the CITS approach to the 20% most Republican-voting localities (hereafter referred to as red states/counties) and to the 20% most Democratic-voting localities (hereafter referred to as blue states/counties) based on 2020 presidential election vote share. Our subgroup analysis comparing CITS effect estimates in blue vs. red states aimed to investigate whether the initial response to imminent safe and effective COVID-19 vaccines showed evidence of heterogeneity by predominant political ideology, under the hypothesis that if conservative voters were more likely to be averse to vaccination, any impact on mental health related symptom search density from the announcement of upcoming vaccines would be lesser in red states and counties compared to blue ones.

@@

The Institutional Review Board (IRB) of the Harvard T.H. Chan School of Public Health determined that the study was not human subjects research, and that additional review was not required (protocol number: IRB23-0095, determined on January 25, 2023).

@@

For most symptoms, search density exhibited negligible differences following the announcement, with small positive level shifts offset by modest negative slope changes. For four mental health symptoms, however, we observed relatively large, negative slope changes for searches for anxiety `r function_dynamic_text_results("its","anxiety", "slope", value_include = "allnumber")`, depression `r function_dynamic_text_results("its","depression", "slope", value_include = "allnumber")`, major depressive disorder (MDD)`r function_dynamic_text_results("its","major_depressive_disorder", "slope", value_include = "allnumber")`, and attention deficit hyperactivity disorder (ADHD)`r function_dynamic_text_results("its","attention_deficit_hyperactivity_disorder", "slope", value_include = "allnumber")`. With respect to level shifts, changes for anxiety `r function_dynamic_text_results("its","anxiety", "intercept", value_include = "allnumber")` and depression `r function_dynamic_text_results("its","depression", "intercept", value_include = "allnumber")` were close to zero, while those for MDD `r function_dynamic_text_results("its","major_depressive_disorder", "intercept", value_include = "allnumber")` and ADHD `r function_dynamic_text_results("its","attention_deficit_hyperactivity_disorder", "intercept", value_include = "allnumber")` were approximately 0.3 percentage points.

@@

While our ITS analysis provided suggestive evidence that search density for mental health symptoms might have changed in the post-announcement period, it also highlighted that searches may have generally changed across other diseases at the same time. Our CITS estimates of each mental health symptom (Figure \ref{figure_cits_outputs}) using searches for headaches as a comparison broadly echoed our ITS analysis, with most conditions showing level and slope changes close to 0. For three conditions, our CITS analysis showed relatively large slope changes with searches for anxiety `r function_dynamic_text_results("main","anxiety", "slope", value_include = "allnumber")`, depression `r function_dynamic_text_results("main","depression", "slope", value_include = "allnumber")`, and MDD `r function_dynamic_text_results("main","major_depressive_disorder", "slope", value_include = "allnumber")`, all showing sustained declines. All three symptoms had relatively large slope changes in our ITS analysis. In addition, the CITS level shift (i.e., the relative immediate effect) for anxiety was `r function_dynamic_text_results("main","anxiety", "intercept")`, a decrease double in magnitude compared to the majority of estimates for other symptoms.

```{r figure_cits_outputs, include = TRUE, echo=FALSE, warnings = FALSE, fig.width = 7, fig.height = 8, fig.align='center', out.width = '95%', fig.scap="Controlled interrupted time series effect estimates on Google Search density by symptom.", out.extra='',  fig.cap="\\label{figure_cits_outputs}Controlled interrupted time series effect estimates (A: slope change per 10 days and B: level shift)  on Google Search density by symptom. Effect estimates are represented as thick, horizontal lines in between colored bands, which represent the 99\\% confidence intervals for the estimates. Colors reflect the sign and magnitude of effect estimates. Dark blue values above zero indicate an increase in search density, while the yellow and green negative values indicate a decrease. The dotted horizontal line depicts a null effect. ADHD: Attention deficit hyperactivity disorder. CoC: clouding of consciousness. GAD: Generalized anxiety disorder. MDD: Major depressive disorder."}

@@

In both our analyses, search trends for anxiety, depression, and MDD showed potential immediate and sustained declines in density following November 16th. Search density for these three symptoms and the comparison search data for headaches is plotted in Figures \ref{figures_cits_data_anxiety} through \ref{figures_cits_data_mdd} at the level of U.S. states, U.S. counties, and international data. We observed relatively strong concordance between our CITS estimates and the observed trends in the data, with strong fit to both mental health symptoms and headaches. We further observed strong within-week patterns in searches, supporting our inclusion of fixed effects to account for this weekly periodicity.

```{r figures_cits_data_anxiety, include = TRUE, echo=FALSE, warnings = FALSE, fig.width = 4.5, fig.height = 7, fig.align='center', out.width = '99%', fig.scap="Density of Google Search trends for anxiety, by region, as a function of time.", out.extra='', fig.cap="\\label{figures_cits_data_anxiety}Density of Google Search trends for anxiety, by region and as a function of time. In each row, search density data for two symptoms are displayed: anxiety in blue and headaches in grey. In the first column, data from U.S. states are shown, where each point represents the search density of a specific symptom on a specific date in a specific state. In the second column, data from U.S. counties are shown; because of the large number of counties reporting data, county-level search data are shown as daily boxplots. In the third column, national data from Australia, the United Kingdom, Ireland, New Zealand, Singapore, and the United States are shown. Each graph shows the density of symptom searches on the y-axis, while time is centered and displayed as days prior to and following November 16, 2020 on the x-axis. The solid lines show predicted lines of best fit under a basic controlled interrupted time series model (i.e., no fixed effects). U.S.: United States."}

@@

Turning to our subgroup analysis, we found modest evidence of effect estimates of similar magnitude comparing across countries (Figure \ref{fig_international}). Across the most populous countries (Australia, the United Kingdom, and the United States), effect estimates were broadly consistent, with wider CIs than our ITS or CITS main analysis results due to the smaller number of observations. Negative level shifts and slope changes were more pronounced in New Zealand compared to other countries, while the level shift estimated for searches for depression in Ireland was positive; we caution, however, that owing to the construction of the Google search density data being relative to regions, this may be due to differences in search volume intensity for depression in Ireland and New Zealand compared to other countries, and not due to a strong effect itself.

>>>>

paper3_abstract <- paste0("For searches related to depression, I observed a sustained decline in search density of ",
             function_dynamic_text_results("main","depression", "slope" , TRUE, TRUE), " and an immediate downward level shift of ",
             function_dynamic_text_results("main","depression", "intercept", TRUE),
             " relative to comparable symptom searches. For anxiety, I estimated a similar slope change of ",
             function_dynamic_text_results("main","anxiety", "slope", TRUE), 
             " and a level shift of ",
             function_dynamic_text_results("main","anxiety", "intercept", TRUE),
 ". I did not observe significant variations across predominately Democrat or Republican voting states and counties, and the findings were robust to alternative specifications of bandwidth, fixed effects, and data cleaning decisions.")

@@

\emph{Daniel Arias}

@@

{\bf Methods:} I utilized controlled interrupted time series approaches to evaluate whether the development of COVID-19 vaccines reduced Google searches related to symptoms of depression and anxiety in U.S. states and counties. I examined how search trends changed following the announcement of vaccine safety and efficacy data for mental health symptoms compared with physical health symptoms, and estimated associations between vaccine announcement and changes in mental health symptom search density. I adjusted for state, month, and weekday effects, and applied numerous sensitivity analyses to assess the robustness of the findings. To explore heterogeneity by predominate political affiliation, I also tested whether the estimates varied among states and counties that predominantly voted for the Democratic Party candidate in the 2020 presidential election compared to the Republican candidate.

{\bf Findings and interpretation:} Under the main model specification, I observed reduced search density for depression and anxiety associated with the announcement of vaccine safety data. `r paper3_abstract` If symptom search density echoes population health, the results would indicate that announcements regarding the development of safe and effective COVID-19 vaccines led to a robust and sustained improvement in mental health. These findings underscore the importance of timely and transparent communication and highlight the potential application of high frequency internet search data for population mental health surveillance.

@@

## Introduction
The COVID-19 pandemic has had a profound effect on the mental health and well-being of individuals worldwide. Studies from Australia,[@batterhamTrajectoriesDepressionAnxiety2021] China,[@ahmedEpidemicCOVID19China2020; @gaoMentalHealthProblems2020; @hwangLonelinessSocialIsolation; @leiComparisonPrevalenceAssociated2020; @liuPrevalencePredictorsPTSS2020; @qiuNationwideSurveyPsychological2020; @wangImmediatePsychologicalResponses2020; @wangStudyPublicPsychological2021; @wangPsychologicalDistressCoping2020; @zhangImpactCOVID19Pandemic2020] Denmark,[@sonderskovDepressiveStateDenmark] Europe,[@hajekPrevalenceDeterminantsProbable2022] Iran,[@moghanibashi-mansouriehAssessingAnxietyLevel2020] Ireland,[@hylandAnxietyDepressionRepublic2020] Italy,[@mazzaNationwideSurveyPsychological2020; @mocciaAffectiveTemperamentAttachment2020] Nepal,[@samadarshiOnlineSurveyFactors2020] New Zealand,[@gasteigerDepressionAnxietyStress2021; @lawRiskFactorsChanges2022] Spain,[@gonzalez-sanguinoMentalHealthConsequences2020; @ozamiz-etxebarriaStressAnxietyDepression2020] Turkey[@ozdinLevelsPredictorsAnxiety2020], the United Kingdom[@dettmannInvestigatingPrevalenceAnxiety2022], and the United States[@jiaNationalStateTrends2021, p. 2020–2021; @kujawaExposureCOVID19Pandemic2020] have reported high levels of anxiety (6-51%), depression (15-48%), and post-traumatic stress disorder (7-54%) associated with the onset of the pandemic. In the United States, approximately four in ten adults reported experiencing symptoms of anxiety or depression in December 2020—a four-fold increase from January-June 2019.[@panchalImplicationsCOVID19Mental2021, p. 19] Globally, it has been estimated that the prevalence of depression and anxiety during the COVID-19 pandemic may be as high as 24% and 21%, respectively.[@banko-ferranImpactCOVID19Mental2020; @castaldelli-maiaInvestigatingEffectNational2021] This striking burden has been described as a mental health crisis[@melimopoulosHowCOVID19Pandemic2021] and as a pandemic in its own right.[@kwaiWhatPointYoung2021]

@@

While this literature suggests that receiving a COVID-19 vaccine can improve mental health, the potential mental health benefits of announcements regarding the development of vaccines have received comparatively less attention.[@cotfasLongestMonthAnalyzing2021; @kesselEffectCOVID19Pandemic2023] In the context of a deadly pandemic, the announcement of a safe and effective vaccine could be a powerful signal, providing a psychological boost that the worst might soon be mitigated. Studying how population mental health may have been impacted by the release of vaccine efficacy and safety trial data may provide insights into how emotional well-being changed throughout the pandemic in response to these signals, which, in turn, could inform public health communication and contribute to the understanding of how the public responds to information about vaccines—a necessary understanding for increasing vaccine uptake and trust in public health responses.

@@

This study analyzes whether news announcements (in November 2020) regarding the imminent availability of safe and effective COVID-19 vaccines impacted population mental health—even before these vaccines were made available to the public. Using Google search density data, I applied interrupted time series approaches to investigate how these announcements might have affected search trends for mental health symptoms in the U.S. and five other countries where data were available: Australia, Ireland, New Zealand, Singapore, and the United Kingdom. Owing to emerging evidence that political ideology may be a predictor of confidence in public health and vaccine hesitancy,[@debusPoliticalIdeologyVaccination2021; @baumgaertnerInfluencePoliticalIdeology2018; @stolerSociopoliticalPsychologicalCorrelates2022] I further explored whether changes in search trends for mental health symptoms varied across U.S. states and counties that predominantly voted Democrat versus Republican in the 2020 presidential election. These findings suggest declines in density for searches related to anxiety and depression in all of the countries examined; in the U.S., these declines were found to be consistent irrespective of political ideology.

@@

I took November 16, 2020 as the date of the first credible signal of the imminent availability of safe and effective COVID-19 vaccines. On November 9^th^, Pzifer announced that early vaccine trial data showed 90% efficacy[@thomasPfizerEarlyData2020] and on the 16^th^, Moderna announced preliminary results showing 95% efficacy.[@gradyEarlyDataShow2020] While U.S. Food and Drug Administration (FDA) Emergency Use Authorization (EUA) was not granted until December, the concurrent announcements of vaccine efficacy were widely reported and interpreted by many public health practitioners, policymakers, and the general public as a substantial signal of a new phase of the pandemic, with Moderna’s chief executive officer calling the news “a game changer” in the fight against COVID-19.

@@

I accessed aggregated national and subnational data from Google’s COVID-19 Search Trends Symptoms Dataset [@Wahltinez2020] (hereafter referred to as the symptoms dataset) between January 1, 2020 and April 1, 2021. The symptoms dataset provides time series data of search term density, standardized to pairs of health symptoms and regions (e.g., searches for depression in Australia or fever in Cook County, Illinois) and normalized within each pair based on the symptom’s relative popularity. The time series of search density for each symptom-region pair also contains artificial noise, as Google’s differential privacy technique to data aggregation and anonymization adds random noise to protect user privacy (the errors are symmetrically distributed).[@Wahltinez2020]

Search density data for over 400 symptoms were reported by Google across national and subnational regions of six predominately English-speaking countries: Australia, Ireland, New Zealand, Singapore, the United Kingdom, and the United States. In the U.S., subnational regions include states and counties. Because the aggregated search density data were constructed based on relative search population within each region, I could not compare search density for specific symptoms across different regions; however, comparing search density within regions (or a group of regions) was feasible both over time and across different symptoms.

@@

I utilized three levels of region: 1) national level for the six countries; 2) U.S. state level data across `r n_states` states and the District of Columbia; and 3) U.S. county level data across `r n_counties` counties (though there are over 3,000 counties in the U.S., due to data privacy concerns, smaller county results were not included in the symptoms dataset). Following the convention of the symptoms dataset, I will refer to the geographic units (e.g., states, counties, and nations) within the data as "regions."

@@

I obtained county-level presidential election results for 2020 from the MIT Election Data and Science Lab,[@mitelectiondataandsciencelabCountyPresidentialElection2022] which I used to calculate aggregate county and state votes (including absentee, early, and election day ballots) for each candidate. The aggregate votes were then used to determine vote share and margins (e.g., the percentage points difference in vote share between the winning candidate and the runner up). Applying a similar approach as other research into COVID-19 and political party inclination,[@grossmannAllStatesClose2021] I used these margins to sort counties and states into quintiles, such that the 20% most Republican-voting localities appearing in the symptoms dataset were sorted into the first quintile and the 20% most Democratic-voting localities sorted into the fifth quintile.

@@

I used an interrupted time series (ITS) approach to model the Google search density for a given symptom $d$ using the following baseline equation: $$Y_{d,g, i} = \lambda_0 + \lambda_1 time_t + \alpha_d Post_{d,i} + \beta_d time_t*post_{d,i} + \theta_r + T^{'}_t\chi + \epsilon_{it} \ \ \ ,\ \ \ \ \ \ \ \ \ \ \ \ (3.1)$$ where $time_t$ is a continuous variable that captures the days before the vaccine announcement ($time_0$), $Post_i$ is an indicator variable equal to 1 in the period after the announcement (0 otherwise), $\theta_r$ indicates a series of fixed effects for each region $(r)$, $T^{'}_t\chi$ is a matrix of time fixed effects (i.e., month and day of week), and $\epsilon_{it}$ is an error term. $\alpha_d$ can be interpreted as a level change associated with the announcement date—that is, the immediate difference in search density for a disease $d$ following November 19, 2020; $\beta_d$—the slope change—can be interpreted as the sustained change in search density in the days following November 16th.

@@

```
I applied this model for all available `r diseases_n` diseases with county-level data. As search density may rapidly change in response to new announcements, I trimmed the bandwidth of dates to fit the model to 40 days before and after November 16, 2020.

To control for time-varying confounders which may affect symptom search trends, I applied a controlled interrupted time series (CITS) approach to contrast how searches for mental health symptoms varied above and beyond changes in search density for a comparable physical complaint (i.e., headaches).

After reviewing the list of symptoms for which search density data were available, I identified the following `r mh_diseases_n` mental health symptoms: `r paste0(gsub("asperger", "Asperger", gsub("_", " ", printList(toPrint = sort(list_mental_health_symptoms), finalSepWord = "and", midSep = ","))),".")`

For each of these symptoms, $d$, I ran the following model:

@@

Given the high number of observations in the sample, I evaluated significance against an alpha of 0.01, and I constructed 99% confidence intervals (CIs) using heteroskedasticity-robust standard errors.

@@

As with other quasi-experimental methods, it is recommended that the modeling specifications be evaluated using a sensitivity analysis. To do so, I defined a list of seven methodological choices concerning data processing and modeling assumptions and interacted these choices to generate all plausible combinations. These combinations were then conducted to allow for a multiverse analysis to identify how sensitive effect estimates and CIs were to different methodological choices. The criteria used to generate the plausible model combinations are outlined in Table 3.1.

@@

First, I considered the role of data cleaning in the outcome data, which involved outlier smoothing using Friedman’s nonparametric super smoother regression estimator.[@friedmanVariableSpanScatterplot1984] Second, I compared the level of region of aggregated data, using U.S. counties, U.S. states, and, finally, national trends from the six counties in the sample. Third, I evaluated sensitivity to the specification of the bandwidth cut-off around the announcement date, extending it from ±40 days to ±50 and ±60 days. Fourth, I toggled whether to account for day of week fixed effects; fifth, I toggled month fixed effects; and sixth, I toggled regional fixed effects (i.e., fixed effects for states in the U.S. state level data, fixed effects for counties in the U.S. county level data, etc.).

@@

The first subgroup analysis applied the CITS approach against all `r mh_diseases_n` mental health related symptoms against search data from each of the six countries in the sample, using the nationally aggregated data. This analysis aimed to investigate the extent to which effects observed in U.S. states and counties were also seen in other countries.

The second subgroup analysis applied the CITS approach to the 20% most Republican-voting localities (hereafter referred to as red states/counties) and to the 20% most Democratic-voting localities (hereafter referred to as blue states/counties) based on 2020 presidential election vote share. The subgroup analysis comparing CITS effect estimates in blue versus red states aimed to investigate whether the initial response to imminent safe and effective COVID-19 vaccines showed evidence of heterogeneity by predominant political ideology, under the hypothesis that if conservative voters were more likely to be averse to vaccination, any impact on mental health related symptom search density from the announcement of upcoming vaccines would be lesser in red states and counties compared to blue ones.

@@

The Institutional Review Board (IRB) of the Harvard T.H. Chan School of Public Health determined that this study was not human subjects research and that additional review was not required (protocol number: IRB23-0095, determined on January 25, 2023).

@@

For most symptoms, search density exhibited negligible differences following the announcement, with small positive level shifts offset by modest negative slope changes. For four mental health symptoms, however, I observed relatively large, negative slope changes for searches for anxiety `r function_dynamic_text_results("its","anxiety", "slope", value_include = "allnumber")`, depression `r function_dynamic_text_results("its","depression", "slope", value_include = "allnumber")`, major depressive disorder (MDD)`r function_dynamic_text_results("its","major_depressive_disorder", "slope", value_include = "allnumber")`, and attention deficit hyperactivity disorder (ADHD)`r function_dynamic_text_results("its","attention_deficit_hyperactivity_disorder", "slope", value_include = "allnumber")`. With respect to level shifts, changes for anxiety `r function_dynamic_text_results("its","anxiety", "intercept", value_include = "allnumber")` and depression `r function_dynamic_text_results("its","depression", "intercept", value_include = "allnumber")` were close to zero, while those for MDD `r function_dynamic_text_results("its","major_depressive_disorder", "intercept", value_include = "allnumber")` and ADHD `r function_dynamic_text_results("its","attention_deficit_hyperactivity_disorder", "intercept", value_include = "allnumber")` were approximately 0.3 percentage points.

@@

While the ITS analysis provided suggestive evidence that search density for mental health symptoms might have changed in the post-announcement period, it also highlighted that searches may have generally changed across other diseases at the same time. The CITS estimates of each mental health symptom (Figure \ref{figure_cits_outputs}) using searches for headaches as a comparison broadly echoed the ITS analysis, with most conditions showing level and slope changes close to 0. For three conditions, the CITS analysis showed relatively large slope changes with searches for anxiety `r function_dynamic_text_results("main","anxiety", "slope", value_include = "allnumber")`, depression `r function_dynamic_text_results("main","depression", "slope", value_include = "allnumber")`, and MDD `r function_dynamic_text_results("main","major_depressive_disorder", "slope", value_include = "allnumber")`, all showing sustained declines. All three symptoms had relatively large slope changes in the ITS analysis. In addition, the CITS level shift (i.e., the relative immediate effect) for anxiety was `r function_dynamic_text_results("main","anxiety", "intercept")`, a decrease double in magnitude compared to most estimates for other symptoms.

```{r figure_cits_outputs, include = TRUE, echo=FALSE, warnings = FALSE, fig.width = 7, fig.height = 8, fig.align='center', out.width = '95%', fig.scap="Controlled interrupted time series effect estimates for Google Search density by symptom.", out.extra='', fig.cap="\\label{figure_cits_outputs}Controlled interrupted time series effect estimates (A: slope change per 10 days and B: level shift) for Google Search density by symptom. Effect estimates are represented as thick, horizontal lines in between colored bands, which represent the 99\\% confidence intervals for the estimates. Colors reflect the sign and magnitude of effect estimates. Dark blue values above zero indicate an increase in search density, while the yellow and green negative values indicate a decrease. The dotted horizontal line depicts a null effect. ADHD: Attention deficit hyperactivity disorder. CoC: clouding of consciousness. GAD: Generalized anxiety disorder. MDD: Major depressive disorder."}

@@

In both the analyses, search trends for anxiety, depression, and MDD showed potential immediate and sustained declines in density following November 16th. Search density for these three symptoms and the comparison search data for headaches is plotted in Figures \ref{figures_cits_data_anxiety} through \ref{figures_cits_data_mdd} at the level of U.S. states, U.S. counties, and international data. I observed relatively strong concordance between the CITS estimates and the observed trends in the data, with strong fit to both mental health symptoms and headaches. I further observed strong within-week patterns in searches, supporting the inclusion of fixed effects to account for this weekly periodicity.

```{r figures_cits_data_anxiety, include = TRUE, echo=FALSE, warnings = FALSE, fig.width = 4.5, fig.height = 6.8, fig.align='center', out.width = '99%', fig.scap="Density of Google Search trends for anxiety, by region, as a function of time.", out.extra='', fig.cap="\\label{figures_cits_data_anxiety}Density of Google Search trends for anxiety, by region and as a function of time. In each row, search density data for two symptoms are displayed: anxiety in blue and headaches in grey. In the first row, data from U.S. states are shown, where each point represents the search density of a specific symptom on a specific date in a specific state. In the second row, data from U.S. counties are shown; because of the large number of counties reporting data, county-level search data are summarized as daily boxplots. In the third row, national data from Australia, Ireland, New Zealand, Singapore, the United Kingdom, and the United States are shown. Each graph shows the density of symptom searches on the y-axis, while time is centered and displayed as days prior to and following November 16, 2020 on the x-axis. The solid lines show predicted lines of best fit under a basic controlled interrupted time series model (i.e., no fixed effects). U.S.: United States."}

@@

Turning to the subgroup analysis, I found modest evidence of effect estimates of similar magnitude comparing across countries (Figure \ref{fig_international}). Across the most populous countries (Australia, the United Kingdom, and the United States), effect estimates were broadly consistent, with wider CIs than the ITS or CITS main analysis results due to the smaller number of observations. Negative level shifts and slope changes were more pronounced in New Zealand compared to other countries, while the level shift estimated for searches for depression in Ireland was positive; owing to the construction of the Google search density data being relative to regions, this may be due to differences in search volume intensity for depression in Ireland and New Zealand compared to other countries, and not due to a strong effect itself.
